---
title: "Lecture 7 Script Template"
author: "Jiaxi Li"
date: "2025-10-21"
output: pdf_document
---

```{r setup, include=FALSE}
# This is global set up for all later chunks
knitr::opts_chunk$set(echo = TRUE)
```

## `tidyverse` Setup

```{r}
# Install the packages if needed. (This takes a while)
install.packages(c('tidyverse', 'nycflights13'), repos = 'https://cran.rstudio.com', dependencies  = TRUE)

# Load the core tidyverse packages
library(tidyverse)
```

```{r}
# View all tidyverse packages
tidyverse_packages()
```

## `dplyr`

#### View data

```{r}
# library(dplyr)
# Note that we will use pipeline extensively from now on
glimpse(starwars)
```

#### filter

```{r}
# filter using logical
starwars |> 
  filter( 
    species == "Human",
    height >= 190
  )
```

```{r}
# base R version:
starbool = starwars$species == "Human" & starwars$height >= 190
starbool

# Note that we need to remove the NA's as well
# You can use replace_na in dplyr!
starwars[?(starbool, F),]
```

```{r}
# Regular expression to keep rows
starwars |> 
  filter(grepl("Skywalker", name))

# Filter out missing data
starwars |> 
  filter(is.na(height))
```

#### arrange

```{r}
# arrange items in ascending order
starwars |> 
  arrange(birth_year)
# arrange items in descending order
starwars |> 
  arrange(desc(birth_year))
```

#### select

```{r}
# select with : and -
starwars |> 
  select(name:skin_color,species,-height)
```

```{r}
# select while change names
starwars |> 
  select(alias=name, crib=homeworld, sex=gender)
starwars |> 
  select(name, homeworld, gender) |>
  rename(alias=name, crib=homeworld, sex=gender)
```

```{r}
# You can also use character vectors
starwars |> 
  select(c("name", "homeworld", "gender"))
```

```{r}
# Use select containing some words
starwars |> 
  select(name, contains("color"))
```

```{r}
# Use select to rearrange variables
starwars |> 
  select(species, homeworld, everything()) |>
  head(5)
```

#### mutate

```{r}
# Use mutate to create a num variable dog_years and character variable comment
starwars |> 
  select(name, birth_year) |>
  mutate(dog_years = birth_year * 7) |>
  mutate(comment = paste0(name, " is ", dog_years, " in dog years."))
```

```{r}
# mutate is order aware
starwars |> 
  select(name, birth_year) |>
  mutate(
    dog_years = birth_year * 7, ## Separate with a comma
    comment = paste0(name, " is ", dog_years, " in dog years.")
  )
```

```{r}
# The first five rows of starwars (for comparison)
starwars |> 
  head(5)

# mutate all character variables (to upper case)
starwars |> 
  select(name:eye_color) |> 
  mutate(across(where(is.character),
                toupper)) |> #<< 
  head(5)
```

```{r}
# The first five rows of starwars (for comparison)
starwars |> 
  head(5)

# mutate all numerical variables with customized functions (from x to x+1)
starwars |> 
  mutate(across(where(?),
                      function(x){return(x+1)})) |>
  head(5)

# # You can also predefine the function and use it in mutate
# add_one = function(x) x+1
# # add_one = function(x) {return(x+1)}
# starwars |> 
#   mutate(across(where(is.numeric), add_one)) |>
#   head(5)
```

#### summarize

```{r}
# summarize with each group of species/gender
starwars |> 
  group_by(species, gender) |> 
  summarize(mean_height = mean(height, na.rm = TRUE),
            .groups = "drop")

# Without na.rm=T?
# Without .groups = "drop"?
starwars |> 
  group_by(species, gender) |> 
  summarize(mean_height = mean(height, na.rm = TRUE)) |>
  ungroup()
```

```{r}
# multiple summarized values 
starwars |> 
  group_by(species, gender) |> 
  summarize(mean_height = mean(height, na.rm = TRUE),
            min_height = min(height, na.rm=TRUE),
            max_height = max(height, na.rm=TRUE))
```

```{r}
# summarize with across
starwars |> 
  group_by(species) |> 
  summarize(across(where(is.numeric), function(x) mean(x, na.rm=T))) |> #<<
  head(5)
```

#### Other useful `dplyr` functions

```{r}
# slice a few rows
starwars |>
  slice(c(1, 5))
```

```{r}
# pull as a vector
starwars |> 
  filter(gender=="feminine") |> 
  pull(height)
```

```{r}
# find unique elements
unique(starwars$gender)
# number of unique/distinct elements
n_distinct(starwars$gender)
```

```{r}
# Find unique elements as a tibble
starwars |>
  distinct(gender)

# Find unique elements with counts as a tibble
starwars |>
  count(gender)
```

## Join with `dplyr`

#### left_join

```{r}
# Load the package for the data demonstration
library(nycflights13)

# Load data
data(flights)
data(planes)
data(airlines)
data(airports)

# Check data
flights |>
  distinct(dest)

head(flights)
head(planes)
head(airlines)
head(airports)
```

```{r}
# first combine flights and planes using a left join
firstjoin = left_join(flights, planes)

# Something wrong, we can check where engines does not exist
firstjoin |>
  filter(is.na(engines)) |>
  nrow()
nrow(firstjoin)
```

```{r}
# The year variables in flights and planes have different meanings!
# Check https://cran.r-project.org/web/packages/nycflights13/nycflights13.pdf
# We should not join by year
left_join(flights,
          planes,
          by="tailnum") |>
  filter(is.na(engines)) |>
  nrow()

# Alternative ways to do it
left_join(flights,
          planes,
          by=join_by(tailnum)) |>
  filter(is.na(engines)) |>
  nrow()
```

```{r}
# Rename year to avoid confusion, better practice
left_join(
  flights |> rename(flight_year = year), 
  planes |> rename(year_built = year),
  by=join_by(tailnum)
  ) |>
  filter(is.na(engines)) |>
  nrow()
```

```{r}
# When joining by one variable but with different names in different data
Tampa_flights = left_join(
  flights, 
  airports,
  by=join_by(dest == faa)
  ) |>
  filter(name == "Tampa Intl")

# The alternative is to rename before joining
Tampa_flights = left_join(
  flights, 
  airports |> rename(dest = faa),
  by=join_by(dest)
) |>
  filter(name == "Tampa Intl")
```

#### Practice questions:

For Tampa Intl airport (Tampa Intl) on Jan 1, 2013, figure out:

Q1. Which carrier flew the most flights into Tampa?

Q2. How many Airbus (AIRBUS) planes did JetBlue (JetBlue Airways) fly into Tampa?

Q3. What manufacturers did Delta (Delta Air Lines Inc.) fly into Tampa?

```{r}
# Create the data for Tampa on Jan 1, 2013
Tampa_2013_01_01 = ?
```

```{r}
# Q1: carrier flew the most flights into tampa? 

```

```{r}
# Q2. How many Airbus (AIRBUS) planes did JetBlue (JetBlue Airways) fly into Tampa Intl? 

```

```{r}
# Q3. What manufacturers did Delta (Delta Air Lines Inc.) fly into Tampa?

```

## `tidyr`

#### pivot_longer

```{r}
# Create some random stock data with X Y Z for pivoting
stocks = data.frame( ## Could use "tibble" instead of "data.frame" if you prefer
  time = as.Date('2009-01-01') + 0:1,
  X = rnorm(2, 0, 1),
  Y = rnorm(2, 0, 2),
  Z = rnorm(2, 0, 4)
)

stock_vol = data.frame(
  X_vol = rep(1,2),
  Y_vol = rep(2,2),
  Z_vol = rep(4,2)
)

# Combine and rename
stocks2 = cbind(stocks,
                stock_vol) |> 
  rename(X_price = X,
         Y_price = Y,
         Z_price = Z)
```

```{r}
# Make the data longer by converting all stocks into one column
tidy_stocks = stocks |> 
  pivot_longer(X:Z,
               names_to = "stock",
               values_to = "returns")

# Another way is to specify the columns to keep using "-"
tidy_stocks = stocks |> 
  pivot_longer(-time,
               names_to = "stock",
               values_to = "returns")
```

```{r}
# pivot_longer and separating
# more info: https://stackoverflow.com/questions/69798752/pivot-longer-for-multiple-sets-having-the-same-names-to
stocks2_long = stocks2 |> 
  pivot_longer(-time, 
               names_sep = "_",
               names_to = c("stock",".value"))
```

#### pivot_wider

```{r}
# Make the data wider by make every stock its own column
tidy_stocks |> 
  pivot_wider(names_from = "stock",
              values_from = "returns")
```

```{r}
# Also can make the data wider by make every date its own column
tidy_stocks |> 
  pivot_wider(names_from = "time",
              values_from = "returns")
```

#### separate

```{r}
economists = data.frame(name = c("Adam.David.Smith", "Paul.David.Samuelson", "Milton.David.Friedman"))
economists

# There are three parts of each name here
economists |>
  separate(name, c("first_name","middle_name", "last_name"))
# gives warning
economists |> 
  separate(name, c("first_name", "last_name"))
```

#### unite

```{r}
# Create gdp data
gdp = data.frame(
  yr = rep(2016, times = 4),
  mnth = rep(1, times = 4),
  dy = 1:4,
  gdp = rnorm(4, mean = 100, sd = 2)
)

# unite year,month,day into date
gdp |>
  unite(date,
        c("yr", "mnth", "dy"),
        sep = "-") |>
  as_tibble() |>
  # convert to date
  mutate(date = ?(date))
```

#### Others: crossing

```{r}
crossing(grade = c("Freshmen","Sophomore","Junior","Senior"),
         subject = c("Econ","Math","English"))
```

```{r}
# Similar to expand.grid (in lec5)
expand.grid(c("Freshmen","Sophomore","Junior","Senior"),
            c("Econ","Math","English"))
```
