---
title: "Lecture 9 Script Template"
author: "Jiaxi Li"
date: "2025-11-04"
output: pdf_document
---

```{r setup, include=FALSE}
# This is global set up for all later chunks
knitr::opts_chunk$set(echo = TRUE)
```

## Package Setup

```{r}
# necessary packages
# install.packges(c("L1pack", "gapminder", "glmnet", "tidyverse", "ggrepel"))
library(tidyverse)
library(ggrepel)  # for non-overlapping labels
set.seed(123)
```

## Model Fitting: Simple Linear Regression

#### Viewing Data

```{r}
library(gapminder)
# Subset the Asia countries in year 2007
gp_subset = gapminder[gapminder$continent=="Asia"&gapminder$year==2007,]
```

```{r}
# Plot the lifeExp against gdpPercap
g = ggplot(gp_subset,aes(x=gdpPercap, y=lifeExp)) + 
  geom_point()
g
```

```{r}
# Plot a fitted line
g + geom_smooth(method='?',se=FALSE)
```

#### Fit a Linear Regression

The plot is relatively linear:

$$LifeExp_i = b_0 + b_1 GDPpc_i + e_i$$

Now, let's select our objective function. MSE is good enough and intuitive so let's use it here!

$$MSE = \frac{1}{N}\sum_i e_i^2$$

Create a function to calculate MSE and option to plot:

```{r}
# Build a MSE function for gp_subset
MSE = function(b, Graph = F) {
  # b is a vector containing b0 and b1
  e = ?
  
  # plot
  if (Graph == T){
    (ggplot(gp_subset,aes(x=gdpPercap, y=lifeExp)) + 
       geom_point() + 
       geom_abline(slope = b[2], intercept = b[1]) +
       labs(title = paste("The MSE is", round(mean(e^2)))) + 
       theme(plot.title = element_text(hjust = 0.5))) |>
      print()
  }
  
  return(mean(e^2))
}
```

#### Try different values

Here, let's try $b_0 = 60$ and $b_1 = 0.0001$. You can set them to different values and try.

```{r}
MSE(b = c(60, 0.0001), Graph = T)
```

#### Optimizing

```{r}
# Optimizing the values
optim(c(60, 0.0001), MSE)
```

#### `lm`

```{r}
# Existing function for linear regression
Result = lm(lifeExp~gdpPercap, data=gp_subset)
summary(Result)
```

#### `lad`

```{r}
# LAD regression
lad_res = lad(lifeExp~gdpPercap, data=gp_subset)
summary(lad_res)
```

## Other Models

#### Multiple Linear Regression

```{r}
# Use multiple variables to explain/predict lifeExp
Result = lm(lifeExp~gdpPercap + year, data=gapminder)
summary(Result)
```

#### Polynomial Regressions

```{r}
# Create new squared variable and run lm
Result = gp_subset |>
  mutate(GDPpc_sq = gdpPercap^2) |>
  lm(lifeExp~gdpPercap + GDPpc_sq, data=_)
summary(Result)
```

There is another way. We can directly add variable in lm formula:

```{r}
# Create new squared variable and run lm
Result = lm(?, data=gp_subset)
summary(Result)
```

#### Log Regression

```{r}
# Create new log variable and run lm
Result = gp_subset |>
  mutate(log_GDPpc = log(gdpPercap)) |>
  lm(lifeExp~log_GDPpc, data=_)
summary(Result)
```

#### Ridge, LASSO and Elastic Net

Note that Elastic Net includes Ridge and LASSO.

```{r}
# Ridge, LASSO and Elastic Net
library(glmnet)
#define response variable
y <- mtcars$hp

#define matrix of predictor variables
x <- data.matrix(mtcars[, c('mpg', 'wt', 'drat', 'qsec')])
```

We can set $\alpha = 0$ for Ridge regression. Check $\lambda = 1, 10, 100, etc.$

```{r}
# Ridge Example
coef(glmnet(x, y, alpha= 0, lambda = 1))
```

We can set $\alpha = 1$ for Ridge regression. Check $\lambda = 1, 10, 100, etc.$

```{r}
# LASSO Example
coef(glmnet(x, y, lambda = 1))
```

#### Logistic Regression

```{r}
# Convert 'am' to factor
mtcars$am <- factor(mtcars$am, labels = c("Automatic", "Manual"))

# Fit logistic regression
# Predict transmission based on mpg and hp
Result <- glm(am ~ mpg + hp, data = mtcars, family = binomial)

# View summary
summary(Result)
```

#### K-Means Clustering

```{r}
# Select two variables
mtcars2 <- mtcars |> select(mpg, hp)

# Scale the data (important for k-means)
mtcars_scaled <- scale(mtcars2)

# Set number of clusters
k <- 3


# Apply k-means clustering
set.seed(123)  # for reproducibility
km <- kmeans(mtcars_scaled, centers = k)

# Add cluster assignments to original data
mtcars2$cluster <- factor(km$cluster)
mtcars2$car <- rownames(mtcars)

# Plot with car names
ggplot(mtcars2, aes(x = mpg, y = hp, color = cluster)) +
  geom_point(size = 4) +
  geom_text_repel(aes(label = car), size = 3) +
  labs(title = paste("k-means clustering of mtcars (k =", k, ")"),
       x = "Miles per Gallon (mpg)", 
       y = "Horsepower (hp)",
       color = "Cluster") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")
```
