---
title: "Lecture 8 Script Template"
author: "Jiaxi Li"
date: "2025-10-28"
output: pdf_document
---

```{r setup, include=FALSE}
# This is global set up for all later chunks
knitr::opts_chunk$set(echo = TRUE)
```

## `data.table` Setup

```{r}
# Install the packages if needed and load them
if (!require(pacman)) install.packages('pacman', 
                                       repos = 'https://cran.rstudio.com')
pacman::p_load(dplyr, data.table, dtplyr, tidyfast, 
               microbenchmark, ggplot2, nycflights13)
```

```{r}
# speed comparison: tidyverse vs. data.table
collapse_dplyr = function() {
  storms %>%
    group_by(name, year, month, day) %>% 
    summarize(wind = mean(wind), pressure = mean(pressure), category = dplyr::first(category))
  }

storms_dt = as.data.table(storms)
collapse_dt = function() {
  storms_dt[, .(wind = mean(wind), pressure = mean(pressure), category = first(category)),
            by = .(name, year, month, day)]
  }

microbenchmark(collapse_dplyr(), collapse_dt(), times = 1)
```

## `data.table` basics

```{r}
# data(starwars, package = "dplyr") ## Optional to bring the DF into the global env
# Summarizing by group with dplyr
starwars |> 
  filter(species=="Human") |> 
  group_by(homeworld) |> 
  summarise(mean_height=mean(height))
```

```{r}
# convert to data.table and save it separately
starwars_dt = as.data.table(starwars)
```

```{r}
# Summarizing by group with data.table
starwars_dt[species=="Human", mean(height), by=gender]
```

## Row manipulation

#### filter

```{r}
# Subset by rows (filter)
starwars_dt[height>190 & species=='Human']
```

#### arrange

```{r}
# Order by rows (arrange)
starwars_dt[order(birth_year)]  ## (temporarily) sort by youngest to oldest
starwars_dt[order(-birth_year)] ## (temporarily) sort by oldest to youngest
```

```{r}
# Check the original data.table. Is the order changed?
starwars_dt
```

```{r}
# in-place change the order
setorder(starwars_dt, birth_year, na.last = TRUE)
starwars_dt[1:5, name:birth_year] ## Only print subset
```

## Column manipulation

#### mutate

```{r}
# Modifying columns := (mutate)
# create a new data.table
DT = data.table(x = 1:2)
# mutate new column (x_sq) and display the data.table
DT[, x_sq ? x^2][]
```

```{r}
# Shallow copy!
DT_copy = DT
# remove a column of the copy by assigning to NULL
DT_copy[, x_sq := ?]
# Check what happened to the original DT!
DT
```

```{r}
# mutate new column (x_sq) again
DT[, x_sq := x^2]
# Deep copy!
DT_copy = copy(DT)
# remove a column of the copy by assigning to NULL
DT_copy[, x_sq := ?]
# Check what happened to the original DT!
DT
```

```{r}
# create a new data.table
DT2 = data.table(a = -2:2, b = LETTERS[1:5])
# In dplyr, you cannot do subassign, but rather assign a new vector 
DT2 |>
  mutate(b = ifelse(a < 0, NA, b))
# In data.table, you can subassign and it is faster
DT2[a < 0, b := NA][]
```

```{r}
# Mutate multiple variables with mutate
DT |>
  mutate(y = 3:4,
         y_name = c("three", "four"))
# Is DT changed by mutate?
DT
```

```{r}
# Mutate multiple variables with :=
DT[, ':=' (y = 3:4,
           y_name = c("three", "four"))]
# Is DT changed by :=?
DT
```

```{r}
# Another way to create multiple variables
DT2[, c("newvar1", "newvar2") := .(a+3,a*2)][]
```

```{r}
# How about dynamic assignment?
DT[, ':=' (z = 5:6, z_sq = z^2)][]
# What can we do?
# Chain the operations instead
DT[, z := 5:6][, z_sq := z^2][] 
```

```{r}
# Pipeline with |> and _ as placeholder
DT |>
  _[, xyz := x+y+z] |>
  _[, xyz_sq := xyz^2] |>
  _[] 
```

#### select

```{r}
## Subsetting on columns (select)
starwars_dt[1:2, c(1:3, 10)]
starwars_dt[, c("name", "height", "mass", "homeworld")] ## Also works
starwars_dt[, list(name, height, mass, homeworld)] ## So does this
starwars_dt[1:2, .(name, height, mass, homeworld)]
```

```{r}
# keep all except a few columns
starwars_dt[, ?c("name", "height")]
```

```{r}
# Change names with setnames (rename)
setnames(starwars_dt, old = c("name", "homeworld"), 
         new = c("alias", "crib"))[1:2,]
# Better change it back, in case we use "name" or "homeworld" on a later slide
setnames(starwars_dt, old = c("alias", "crib"), 
         new = c("name", "homeworld"))[1:2,]
```

```{r}
# (temporally) subsetting and renaming at the same time
starwars_dt[1:2, .(alias = name, crib = homeworld)]
# Is the original data modified?
starwars_dt[1:2,]
```

This subsetting would not change the original data. (To remove columns, we need ":= NULL".)

```{r}
# data.table and dplyr used together
starwars_dt[1:5, ] |> 
  select(crib = homeworld, everything())
```

#### summarize

```{r}
## Aggregating (summarize)
# Note that this is copy and modify, the original starwars_dt is not modified
starwars_dt[, mean(height, na.rm=T), by = species]
starwars_dt[, .(species_height = mean(height, na.rm=T)), by = species]
```

```{r}
# Create a new column in original starwars_dt storing the mean_height
starwars_dt[, mean_height := mean(height, na.rm=T)] |> ## Add mean height as column
  _[1:5, .(name, height, mean_height)] ## Just to keep everything on the slide
```

```{r}
# Check the numbers of observations
starwars_dt[, ?]
```

#### Group

```{r}
# Note that in aggregating, we already used by argument
starwars_dt[, .(mean_height = mean(height, na.rm=T)), by = .(species, homeworld)] |>
  head(4)
```

```{r}
# Efficient subsetting with .SD
starwars_dt[, 
            lapply(.SD, mean, na.rm=T),
            .SDcols = height:mass, #c("height", "mass", "birth_year"),
            by = species] |> 
  head(2) ## Just keep everything on the slide
```

```{r}
# Use lapply to modify all columns
DT <- DT[, x:x_sq]
DT[, lapply(.SD, mean)]
```

#### Key

```{r}
## First create a keyed version of the storms data.table.
## Note that key variables match the 'by' grouping variables below.
storms_dt_key = as.data.table(storms, key = c("name", "year", "month", "day"))
# Same dplyr function as before
collapse_dplyr = function() {
  storms |>
    group_by(name, year, month, day) |> 
    summarize(wind = mean(wind), pressure = mean(pressure), category = dplyr::first(category))
}
## Collapse function for this keyed data.table. Everything else stays the same.
collapse_dt_key = function() {
  storms_dt_key[, .(wind = mean(wind), pressure = mean(pressure), category = first(category)), 
                by = .(name, year, month, day)]
}
## Run the benchmark on all three functions.
microbenchmark(collapse_dplyr(), collapse_dt(), collapse_dt_key(), times = 1)
```

## Merging datasets

#### left_join

```{r}
# library(nycflights13) ## Already loaded
flights_dt = as.data.table(flights) 
planes_dt = as.data.table(planes)
airlines_dt = as.data.table(airlines)
airports_dt = as.data.table(airports)
```

```{r}
# dplyr with left_join
left_join(
  flights, 
  planes, 
  by = "tailnum"
)

# data.table with merge
merge(
  flights_dt, 
  planes_dt, 
  all.x = TRUE, ## omit for inner join
  by = "tailnum")
```

```{r}
# We are encountering the name conflicts again
# Need to change variable name "year" from at least one dataset
setnames(planes_dt, old = "year", new = "year_built")
merge(
  flights_dt,
  planes_dt,
  all.x = TRUE, 
  by = "tailnum")
```

```{r}
# There is a slight difference between left_join and merge
# Merge will reorder the left data by the key
# Left_join will not reorder data
# If you want to achieve merge but no reordering, you need option "sort = F"
# setnames(planes_dt, old = "year", new = "year_built")
merge(
  flights_dt,
  planes_dt,
  all.x = TRUE, 
  by = "tailnum",
  ?)
```

#### Practice questions:

For Tampa Intl airport (Tampa Intl) on Jan 1, 2013, figure out:

Q1. Which carrier flew the most flights into Tampa?

Q2. How many Airbus (AIRBUS) planes did JetBlue (JetBlue Airways) fly into Tampa?

Q3. What manufacturers did Delta (Delta Air Lines Inc.) fly into Tampa?

```{r}
# Create the data for Tampa on Jan 1, 2013
setnames(airlines_dt, old = "name", new = "carrier_name")
Tampa_dt_2013_01_01 =
```

```{r}
# Q1: carrier flew the most flights into tampa? 

```

```{r}
# Q2. How many Airbus (AIRBUS) planes did JetBlue (JetBlue Airways) fly into Tampa Intl? 

```

```{r}
# Q3. What manufacturers did Delta (Delta Air Lines Inc.) fly into Tampa?

```

## Reshaping data

#### pivot_longer

```{r}
# Create some random stock data with X Y Z for pivoting
stocks = data.table(time = as.Date('2009-01-01') + 0:1,
                    X = rnorm(2, 0, 1),
                    Y = rnorm(2, 0, 2),
                    Z = rnorm(2, 0, 4))
```

```{r}
## Reshape from wide to long
melt(stocks, id.vars ="?")
```

```{r}
## Reshape from wide to long with defined names
# Use melt to pivot_wider
stocks_long = melt(stocks, id.vars ="time", 
                   variable.name = "stock", value.name = "price")
stocks_long
```

```{r}
# Use dt_pivot_longer to pivot_wider
stocks |> 
  dt_pivot_longer(X:Z, names_to="stock", values_to="price")
```

#### pivot_wider

```{r}
## Reshape from long to wide
# Use dcast to pivot_longer
dcast(stocks_long, 
      time ~ stock, 
      value.var = "price")
```

```{r}
# Use dt_pivot_wider to pivot_longer
stocks_long |> 
  dt_pivot_wider(names_from=stock, 
                 values_from=price)
```

## use `dtplyr` package lazy_dt

```{r}
# library(dtplyr)
# convert data to lazy_dt and then use dplyr syntax
lazy_dt(starwars) |>
  filter(species=="Human") |>
  group_by(gender) |>
  summarise(height = mean(height, na.rm=TRUE))
```
