---
title: "Sentiment Analysis with Ollamar"
author: "Jiaxi Li"
date: "2025-12-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The following code is from: <https://mlverse.github.io/mall/>.

```{r}
# install.packages("ollamar")
library(ollamar)

test_connection()  # test connection to Ollama server
# if you see "Ollama local server not running or wrong server," Ollama app/server isn't running

# # download a model
# # This code only needs to run once and we can delete unuseful model using delete function
# pull("llama3.2")  # download a model (equivalent bash code: ollama run llama3.2)

# generate a response/text based on a prompt; returns an httr2 response by default
resp <- generate("llama3.2", "tell me a 5-word story")
resp

#' interpret httr2 response object
#' <httr2_response>
#' POST http://127.0.0.1:11434/api/generate  # endpoint
#' Status: 200 OK  # if successful, status code should be 200 OK
#' Content-Type: application/json
#' Body: In memory (414 bytes)

# get just the text from the response object
resp_process(resp, "text")
# get the text as a tibble dataframe
resp_process(resp, "df")

# alternatively, specify the output type when calling the function initially
txt <- generate("llama3.2", "tell me a 5-word story", output = "text")

# list available models (models you've pulled/downloaded)
list_models()
```

## Sentiment Analysis

Now, apply the sentiment analysis. `mall` package contains reviews from

```{r}
library(mall)

data("reviews")

reviews <- reviews |>
  rbind("I have no idea about it.")
```

```{r}
reviews |>
  llm_sentiment(review)
```

We can also assign values to each sentiment category.

```{r}
reviews |>
  llm_sentiment(review, c("positive" ~ 1, "neutral" ~ 0.5, "negative" ~ 0))
```

Something is off here between the number and category though. You can decide which to use, but the developer result for the third statement is neutral.

We can further ask about a sentiment score.

```{r}
reviews |>
  llm_sentiment(review, c("positive" ~ 1, "neutral" ~ 0.5, "negative" ~ 0), additional_prompt = "")
```

You can evaluate a vector instead of dataframe.

```{r}
llm_vec_sentiment(c("I am happy", "I am sad"), c("positive" ~ 1, "neutral" ~ 0.5, "negative" ~ 0))
```

You can look into the prompt that is given to Ollama.

```{r}
llm_vec_sentiment(c("I am happy", "I am sad"), c("positive" ~ 1, "neutral" ~ 0.5, "negative" ~ 0), preview = TRUE)
```

It can be used for other purposes as well! Check out the <https://mlverse.github.io/mall/>.
