---
title: "Torch with MINST Digits"
author: "Jiaxi Li"
date: "2025-12-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Hand-written digits recognition is a classical example of Neural Networks applications.

This example and code are from: <https://hastie.su.domains/ISLR2/Labs/R_Labs/Ch10-deeplearning-lab-torch.R>

There are adjustments and comments by ChatGPT and me.

## Setup

#### Packages

```{r}
# Set a garbage collector
options(torch.threshold_call_gc = 4000)
library(torch)
library(luz) # high-level interface for torch
library(torchvision) # for datasets and image transformation
library(torchdatasets) # for datasets we are going to use
torch_manual_seed(13)
```

#### Download data and check dimensions

```{r}
# Load MNIST dataset (train and test)
train_ds <- mnist_dataset(root = ".", train = TRUE, download = TRUE)
test_ds  <- mnist_dataset(root = ".", train = FALSE, download = TRUE)

# Inspect the structure of a single observation
str(train_ds[1])
str(test_ds[2])

# Check dataset sizes
length(train_ds)
length(test_ds)
```

#### Reshape Data

```{r}
# Define a transform function: convert to tensor, flatten, scale 0–1
transform <- function(x) {
  x |>
    torch_tensor(dtype = torch_float()) |>
    torch_flatten() |>
    torch_div(255)
}

# Reload datasets using the transform
train_ds <- mnist_dataset(
  root = ".",
  train = TRUE,
  download = TRUE,
  transform = transform
)
test_ds <- mnist_dataset(
  root = ".",
  train = FALSE,
  download = TRUE,
  transform = transform
)

# Inspect the structure of a single observation
str(train_ds[1])
str(test_ds[2])
```

## 2 Layer Model

#### Model Setup

```{r}
# Define a neural network with 2 hidden layers + dropout
modelnn <- nn_module(
  initialize = function() {

    # Fully connected layers
    self$linear1 <- nn_linear(in_features = 28*28, out_features = 256)
    self$linear2 <- nn_linear(in_features = 256, out_features = 128)
    self$output <- nn_linear(in_features = 128, out_features = 10)

    # Dropout layers
    self$drop1 <- nn_dropout(p = 0.4)
    self$drop2 <- nn_dropout(p = 0.3)

    # Activation
    self$activation <- nn_relu()
  },

  forward = function(x) {
    # Reshape to ... x 784
    x <- x$view(c(784, -1))$t()  # -1 automatically infers the correct number of columns and then transpose
    
    x |>
      self$linear1() |>
      self$activation() |>
      self$drop1() |>
      self$linear2() |>
      self$activation() |>
      self$drop2() |>
      self$output()
  }
)
```

#### Architechture

```{r}
# Print model architecture
print(modelnn())
```

#### Loss Function and Optimizer

```{r}
# Attach loss, optimizer, and metric
modelnn <- modelnn |>
  setup(
    loss = nn_cross_entropy_loss(),
    optimizer = optim_rmsprop,
    metrics = list(luz_metric_accuracy())
  )
```

#### Fit Model

```{r}
# Train model for 5 epochs, 20% validation split
system.time(
  fitted <- modelnn |>
    fit(
      data = train_ds,
      epochs = 5,
      valid_data = 0.2,
      dataloader_options = list(batch_size = 256),
      verbose = FALSE
    )
)

# Plot training/validation performance
plot(fitted)
```

#### Evaluate

```{r}
# Predict classes and compute accuracy
pred_NN <- fitted |>
  predict(test_ds) |>
  torch_argmax(dim = 2) |> 
  as_array()

# Check accuracy with Confusion Matrix, make the threshold to be 0.5
prop.table(table(Predicted = pred_NN,
                 Actual = test_ds$targets))
```

```{r}
# Accuracy helper function
accuracy <- function(pred, truth) {
  mean(pred == truth)
}

# True labels from the test set
truth <- sapply(seq_along(test_ds), function(x) test_ds[x][[2]])

# Use the function for accuracy
accuracy(pred_NN, truth)
```

## 0 Layer Model: Logistic

#### Model Setup

```{r}
# A simple logistic regression model (output layer only)
modellr <- nn_module(
  initialize = function() {
    self$output <- nn_linear(784, 10)
  },
  forward = function(x) {
    # Reshape to ... x 784
    x <- x$view(c(784, -1))$t()  # -1 automatically infers the correct number of columns and then transpose
    
    x |>
      self$output()
  }
)

print(modellr())
```

#### Fit Model

```{r}
# Train logistic regression model
fit_modellr <- modellr |>
  setup(
    loss = nn_cross_entropy_loss(),
    optimizer = optim_rmsprop,
    metrics = list(luz_metric_accuracy())
  ) |>
  fit(
    data = train_ds,
    epochs = 5,
    valid_data = 0.2,
    dataloader_options = list(batch_size = 128)
  )
```

#### Evaluate

```{r}
# Evaluate accuracy on test data
pred_lr <- fit_modellr |>
  predict(test_ds) |>
  torch_argmax(dim = 2) |> 
  as_array()

# Check accuracy with Confusion Matrix, make the threshold to be 0.5
prop.table(table(Predicted = pred_lr,
                 Actual = test_ds$targets))
```

```{r}
# Estimate accuracy
accuracy(pred_lr, truth)

# Or use the built-in evaluate() function for accuracy
evaluate(fit_modellr, test_ds)
```
