---
Ttitle: "Torch Sentiment"
author: "Jiaxi Li"
date: "2025-12-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a example sentiment analysis with IMDB Movie Review data. You can check other types of data (like mltrev23 / Financial Sentiment Analysis) to analyze other types of sentiment.

Example and code are from: <https://hastie.su.domains/ISLR2/Labs/R_Labs/Ch10-deeplearning-lab-torch.R>

There are adjustments and comments by ChatGPT and me.

## Setup

#### Packages

```{r}
# Set a garbage collector
options(torch.threshold_call_gc = 4000)
library(torch)
library(luz)
library(torchdatasets) # for datasets we are going to use
library(Matrix)
library(glmnet)
library(purrr)
```

#### Download data and check

```{r}
# Load the IMDb dataset (max_features = 10000)
max_features <- 10000
imdb_train <- imdb_dataset(root = ".", download = TRUE, num_words = max_features)
imdb_test <- imdb_dataset(root = ".", download = TRUE, num_words = max_features)

# Inspect the first few tokens of a training sample
imdb_train[1]$x[1:12]
```

#### Decoder to words

```{r}
# Decode review into words
word_index <- imdb_train$vocabulary
decode_review <- function(text, word_index) {
  word <- names(word_index)
  idx <- unlist(word_index, use.names = FALSE)
  word <- c("<PAD>", "<START>", "<UNK>", word)
  words <- word[text]
  paste(words, collapse = " ")
}
decode_review(imdb_train[1]$x[1:12], word_index)
```

#### Encoding Function

```{r}
# One-hot (categorical variable) encoding function
one_hot <- function(sequences, dimension) {
  seqlen <- sapply(sequences, length)
  n <- length(seqlen)
  rowind <- rep(1:n, seqlen)
  colind <- unlist(sequences)
  sparseMatrix(i = rowind, j = colind, dims = c(n, dimension))
}
```

#### Data Splitting

```{r}
# Collect all training and test data into lists
train <- seq_along(imdb_train) |>
  lapply(function(i) imdb_train[i]) |>
  purrr::transpose()
test <- seq_along(imdb_test) |>
  lapply(function(i) imdb_test[i]) |>
  purrr::transpose()
```

```{r}
# One-hot encode sequences
x_train_1h <- one_hot(train$x, max_features + 3)  # add 3 for padding/start/oov
x_test_1h <- one_hot(test$x, max_features + 3)

# Check sparsity
dim(x_train_1h)
nnzero(x_train_1h) / (length(train$y) * (max_features + 3))
```

```{r}
# Split validation set
set.seed(2025)
ival <- sample(seq(along = train$y), 2000)
itrain <- seq_along(train$y)[-ival]
```

## Logistic Regression

#### CV

```{r}
# Fit logistic regression on one-hot training data
y_train <- unlist(train$y)
fitlm <- glmnet(x_train_1h[itrain, ], y_train[itrain],
                family = "binomial", standardize = FALSE)

# Predict on validation set and compute accuracy
classlmv <- predict(fitlm, x_train_1h[ival, ]) > 0
acclmv <- apply(classlmv, 2, function(pred) mean(pred == (y_train[ival] > 0)))

# Plot validation accuracy across lambdas
par(mar = c(4, 4, 4, 4), mfrow = c(1, 1))
plot(-log(fitlm$lambda), acclmv)
```

#### Fit and Test

```{r}
# Refit the model with train + validation
# Best lambda
lambda_best <- fitlm$lambda[which.max(acclmv)]
fitlm_best <- glmnet(x_train_1h,
                     y_train,
                     family = "binomial",
                     standardize = FALSE,
                     lambda = lambda_best)

# Test with test set
Test_fit <- predict(fitlm_best, x_test_1h, type = "response")

# Check accuracy with Confusion Matrix, make the threshold to be 0.5
prop.table(table(Predicted = ifelse(Test_fit>0.5, "Positive", "Negative"),
                 Actual = ifelse(as.numeric(test$y), "Positive", "Negative")))
```

```{r}
# Check total accuracy
mean((Test_fit>0.5) == as.numeric(test$y))
```

## Neural Nets

#### Model Setup

```{r}
# Define a small neural network for classification
model <- nn_module(
  initialize = function(input_size = max_features + 3) {
    self$dense1 <- nn_linear(input_size, 16)
    self$relu <- nn_relu()
    self$dense2 <- nn_linear(16, 16)
    self$output <- nn_linear(16, 1)
  },
  forward = function(x) {
    x |>
      self$dense1() |>
      self$relu() |>
      self$dense2() |>
      self$relu() |>
      self$output() |>
      torch_flatten(start_dim = 1)
  }
)
```

#### Loss Function and Optimizer

```{r}
# Set up model with loss, optimizer, and metrics
model <- model |>
  setup(
    loss = nn_bce_with_logits_loss(),
    optimizer = optim_rmsprop,
    metrics = list(luz_metric_binary_accuracy_with_logits())
  ) |>
  set_opt_hparams(lr = 0.001)
```

#### Fit Model as CV

```{r}
# Fit the model on training and check validation set
fitted <- model |>
  fit(
    list(
      torch_tensor(as.matrix(x_train_1h[itrain,]), dtype = torch_float()),
      torch_tensor(y_train[itrain])
    ),
    valid_data = list(
      torch_tensor(as.matrix(x_train_1h[ival, ]), dtype = torch_float()),
      torch_tensor(y_train[ival])
    ),
    dataloader_options = list(batch_size = 512),
    epochs = 10
  )

# Plot training/validation metrics
plot(fitted)
```

#### Fit Model

```{r}
# Check the stopping time
valid_metrics <- do.call(rbind, lapply(fitted$records$metrics$valid, as.data.frame))

# Find the row with best accuracy
Best_epoch <- which.max(valid_metrics$acc)

# Fit the model with Train+Validation and evaluate on test set
fitted_OOS <- model |>
  fit(
    list(
      torch_tensor(as.matrix(x_train_1h), dtype = torch_float()),
      torch_tensor(y_train)
    ),
    valid_data = list(
      torch_tensor(as.matrix(x_test_1h), dtype = torch_float()),
      torch_tensor(unlist(test$y))
    ),
    dataloader_options = list(batch_size = 512),
    epochs = Best_epoch
  )
```

#### Evaluate

```{r}
# Extract the fitted value
# Convert test data to torch tensor
x_test_tensor <- torch_tensor(as.matrix(x_test_1h), dtype = torch_float())

# Get predicted logits (before sigmoid)
pred_logits <- predict(fitted_OOS, x_test_tensor)

# Convert to probabilities (sigmoid)
pred_probs <- torch_sigmoid(pred_logits) |> as_array()

# Check accuracy with Confusion Matrix, make the threshold to be 0.5
prop.table(table(Predicted = ifelse(pred_probs>0.5, "Positive", "Negative"),
                 Actual = ifelse(as.numeric(test$y), "Positive", "Negative")))
```

```{r}
# Check total accuracy
mean((pred_probs>0.5) == as.numeric(test$y))
```

## Test Your Review

Write your own movie review here and check whether the model think it is a good or bad review:

```{r}
# Example review
review <- "This movie is too interesting to keep me awake. What a waste of money!"

# Keep only letters and spaces, convert to lowercase
clean_review <- gsub("[^a-zA-Z ]", "", review)
clean_review <- tolower(clean_review)

# Tokenize text: split by space
tokens <- strsplit(tolower(clean_review), "\\s+")[[1]]

# Map to indices safely
# 1 = <PAD>, 2 = <START>, 3 = <UNK>
indices <- sapply(tokens, function(w) {
  if (w %in% names(word_index)) {
    which(names(word_index) == w)
  } else {
    3   # unknown token
  }
})


# 3. One-hot encode
one_hot_review <- sparseMatrix(
  i = rep(1, length(indices)),
  j = as.integer(indices),
  dims = c(1, max_features + 3) # same as training dim
)

# Convert to torch tensor
x_review <- torch_tensor(as.matrix(one_hot_review), dtype = torch_float())

# 4. Predict with logit and NN
pred_logit <- predict(fitlm_best, one_hot_review, type = "response")
pred_NN <- predict(fitted_OOS, x_review) |>
  torch_sigmoid() |>
  as_array()


cat("Predicted class for Logit:", pred_logit > 0.5, "Probability of positive:", pred_logit, "\n")
cat("Predicted class for NN:", pred_NN > 0.5, "Probability of positive:", pred_NN, "\n")
```
