---
title: "ECON 370: Assignment 3"
author: "Jiaxi Li"
date: "Due: Before class on October 30th, 2025"
output: 
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{geometry}
  - \usepackage{float}
  - \usepackage{graphicx}
  - \usepackage{caption}
  - \usepackage{multirow}
  - \usepackage{array}
  - \usepackage[affil-it]{authblk}
  - \usepackage{lscape}
  - \usepackage[bottom]{footmisc}
  - \usepackage{amssymb}
  - \usepackage{amsmath}
  - \geometry{top=1in, bottom=1in}
  - \setlength{\parindent}{0pt}
  - \setlength{\parskip}{12pt}
---

```{r setup, include=FALSE}
# Note, this is global setting for all code chunks
knitr::opts_chunk$set(eval = FALSE, include = FALSE)
```

# Directions

## As a reminder, LLMs and AIs are allowed from use on this assignment.

This assignment serves dual purpose: **application of `tidyverse`** and **cleaning and merging some of your data for the final project**. You can check **final project details** as a guide. You can use AI but consider using StackOverflow before asking AI for a solution!

Use the Rmd template and write code to answer the following questions. In order to ensure smooth grading, I will ask you to write both your code and explanation (as comment) in code chunk. Please write your code and answer in the corresponding place.

Each question is worth three points. For each section, a "question" corresponds to a number.

Please **comment** your code clearly. 0.5 point per question will be deducted if there is a significantly lack of commenting. Also, the easier your code is to read, the more likely I will be able to figure out what you're doing.

Remember to **host your data on GitHub** if it cannot directly be pulled using functions such as `tq_get`.

There are many ways to clean data. However, I will ask you to apply **pipeline and `tidyverse`** **package** for data wrangling here. 0.5 point per question will be deducted if other method is used while getting the correct answer.

Remember to ***run Section O*** before you proceed. After finishing all code, make sure to **restart** R session and **clear** everything from your workspace. **Rerun** everything to ensure there is no dependency issue in your code.

You need to work individually for this homework.

------------------------------------------------------------------------

# O. Return Data

***This is the code to extract return data online. Do not modify.***

Note: we are using the S&P 500 index returns, not individual companies in the S&P 500.

```{r}
# Data Loading in Section, do not modify
library(tidyquant)
library(tidyverse) # ggplot2 is already in tidyverse
library(lubridate)

# Get S&P500 daily data
SP500_simple_returns = tq_get("^GSPC",
                               from = "1984-01-01",
                               to   = "2024-12-31",
                               get  = "stock.prices") |>
  tq_transmute(
    select     = adjusted,       # adjusted price for accuracy
    mutate_fun = periodReturn,   # calculate period returns
    period     = "monthly",      # monthly returns
    type       = "arithmetic",   # simple return (percentage change)
    col_rename = "monthly_return"
  ) |>
  # Convert to percentage
  mutate(monthly_return = monthly_return*100)

head(SP500_simple_returns)
```

------------------------------------------------------------------------

# I. Variables and Reasoning

In this assignment, I will ask you to pull **three variables each from a different source** (you will need to do at least 10 from at least 4 sources in the final project).

1.  Name the three variables, sources and frequency of your data.

```{r}
# Answer:
# Variable 1:
# Source:
# Frequency:

# Variable 2:
# Source:
# Frequency:

# Variable 3:
# Source:
# Frequency:
```

2.  Reason to include each variable(articles or strong logic support).

```{r}
# Answer:
# Reason to include Variable 1:

# Reason to include Variable 2:

# Reason to include Variable 3:

```

3.  Upload all your **raw** data on GitHub unless they can be extracted using R functions (without downloading things). Make sure they have **long enough history**. Then download each data as a **tibble**:

```{r}
# Code Below, remember commenting

```

------------------------------------------------------------------------

# II. Data Cleaing

1.  Check the **date format** of your datasets. Do you see different formats than "yyyy-mm-dd"? Do you different ways of documenting with month end? If so, fix them with **`mutate`** function combined with `lubridate` package.

```{r}
# Answer about date issue:


# Code Below, remember commenting

```

2.  Use `left_join` to **join** all your data with the return data "SP500_simple_returns".

```{r}
# Code Below, remember commenting

```

3.  What is the situation of **mixed frequency** and **missing data**? If there are, fill in with appropriate method. One reference of mixed frequency is [MIDAS](https://rady.ucsd.edu/_files/faculty-research/valkanov/midas-touch.pdf).

```{r}
# Answer about mixed frequency and missing data issue:


# Code Below, remember commenting

```

4.  What is the situation of **look-ahead bias due to publication lags** for all three variables? Are there any? If there are, use `mutate` and `lag` to make appropriate lags.

```{r}
# Answer about publication lags issue:


# Code Below, remember commenting

```

5.  Use `lead` to create predictive date and predictive return.

```{r}
# Code Below, remember commenting

```

6.  Filtering the desired dates: predictive date from 1985 to 2024.

```{r}
# Code Below, remember commenting

```

------------------------------------------------------------------------

# III. Exploration with `tidyverse`

1.  Make the data longer using `pivot_longer`, each variable name are stored in a column called "Var_name" and all values are stored in a column called "Var_value".

```{r}
# Code Below, remember commenting

```

2.  Use `group_by` to group by different values in "Var_name" and estimate the **mean** for each variable using `summarise`. Use `round` to round to 2 decimal places.

```{r}
# Code Below, remember commenting

```

3.  Use **ggplot** to visualize all the series together in a single plot with each series represented in a different color. Examine (with the plot) whether the series fluctuate around a stable level or display a trend that drifts away over time.

```{r}
# Code Below, remember commenting

```

```{r}
# Answer:

```

Note: if you find one of your variable displays a trend that drifts away over time, you should reconsider how to use that variable. (This is called non-stationary series in time series.)
