---
title: "Data Science for Economists"
# subtitle: "<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>"
subtitle: "Lecture 2: R language basics"
author: "Jiaxi Li"
date: "University of North Carolina | [ECON 370](https://github.com/JiaxiLi1995/ECON370)" #"`r format(Sys.time(), '%d %B %Y')`"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
name: toc

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
  fig.align="center", #fig.width=6, fig.height=4.5, 
  # out.width="748px", #out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=F#, echo=F, warning=F, message=F
  )
library(hrbrthemes)
library(fontawesome)

```

# Table of contents

1. [Prologue](#prologue)

2. [Introduction](#intro)

3. [Types of Data](#datatypes)

4. [Data Strctures](#datastructures)

7. [Getting Help](#help)

`r fa('star-of-life')` Slides adapted from Drew Van Kuiken and Alex Marsh's ECON 370 and originally based on materials by Grant McDermott's EC 607 at University of Oregon.

---
class: inverse, center, middle
name: prologue

# Prologue

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Agenda

Today and the next lecture are going to be very hands on. 

- I'll have slides as per usual, but we're going to spent a lot of time live coding together.

This is deliberate. 

- I want you to get comfortable typing R commands yourself — and navigating the RStudio IDE — without resorting to copy+paste.
- Slightly more painful in the beginning, but much better payoff in the long-run.

---
class: inverse, center, middle
name: intro

# Introduction

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

(Some important R concepts)

---

# Basic Arithmetic

R is a powerful calculator and recognizes all of the standard arithmetic operators:

```{r}
1+2 ## Addition
6-7 ## Subtraction
5/2 ## Division
2^3 ## Exponentiation
2+4*1^3 ## Standard order of precedence (`*` before `+`, etc.)
```

---

# Basic Arithmetic

When possible, do operators in vectors in `R`.

```{r}
first_vec = 1:5  #store our first vector
first_vec        #show vector
first_vec + 0.5  #add 1/2 to each element 
first_vec + 6:10 #add another vector to first_vector
first_vec + 6:9  #oops!
```


---

# Basic Arithmetic (cont.)

We can also invoke modulo operators (quotient & remainder). 
- Very useful when dealing with time, for example.

```{r}
100 %/% 60 ## How many whole hours in 100 minutes?
120 %/% 60 ## How many whole hours in 120 minutes?
100 %% 60 ## How many minutes are left over from dividing 100 by 60?
120 %% 60 ## How many minutes are left over from dividing 120 by 60?
```


---

# More about Vector Operation

`:` sequence: start from the first number, add 1 each time, stop right before become greater than the later number.

```{r}
1.4:3.5
```

</br>

You can also use `c`, concatenation to create a vector (more about `c` later):

```{r}
c(1,2,5,12,1.4)
c("Blue", "Beige", "Black", "Brown")
```


---

# More about Vector Operation

In vector operation, we can have number first and then vector:

```{r}
2^(1:4)
1 + 2:4
```

</br>


Modulo operators can also be applied to vectors:

```{r}
110:130 %% 60
120 %% 15:20
```



---
class: inverse, center, middle
name: datatypes

# Data Types in R

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>


---

# Data Types

R has 6 basic data types:

1. Character
2. Numeric
3. Integer
4. Logical
5. Complex
6. Raw (we will mostly ignore this type)

---

# Data Types (Character)

- The character type can be described as "text."
  - It is known as a "string" in other programming languages.

```{r}
my_name    = "Jiaxi Li"
first_name = "Jiaxi"
last_name  = "Li"     #good style to line up to equals signs (i often don't put in the effort)
class(my_name)
is.character(my_name)

```

---

# Data Types (Character)

## Counting Characters

```{r}
length(my_name) # length also works for other data types
length(c(first_name,last_name))
nchar(my_name)
nchar(c(first_name,last_name))
```

---

# Data Types (Character)

## Comparing (In general)

**Important**: `==` is used for comparing two objects and `=` is used for assigning values.

```{r}
also_my_name = 'Jiaxi Li' # We are using ' instead of " here
my_name == also_my_name

Lowercase_my_name = 'jiaxi li'
my_name == Lowercase_my_name
```

---

# Data Types (Character)

## Combining Characters

```{r,error=TRUE}
paste(first_name,last_name)
paste(first_name,last_name,sep="-")
paste0(first_name,last_name)
first_name + " " + last_name
```

Note: `c` and `paste` are different and `+` does not work for character. We can use `cat` as a combination of `paste` and `print`.

---

# Data Types (Character)

## Printing (In general)

- The following code will look *very* similar to a user when examined in the console.

```{r}
my_name
print(my_name)
```
--

- However, what's happening is ***very different.***
  - The first is called auto-printing, and it only happens in interactive mode (e.g., console or top-level script execution).
  - The second forces printing even in non-interactive contexts (loops and functions). We will talk about loops and functions later in more details.

---

# Data Types (Character)

## Printing (In general)

```{r}
# Things inside of functions/curly brackets are considered non-interactive contexts
nchar(my_name)
nchar(print(my_name))
for (j in 1) {my_name}
for (j in 1) {print(my_name)}
```

---

# Data Types (Character)

## Change Cases

```{r}
toupper(paste(first_name,last_name))
tolower(paste(first_name,last_name))

toupper(c(first_name,last_name))
tolower(c(first_name,last_name))
```

---

# Data Types (Character)

## Special Characters

- Special characters "not native" to English can still be used but must be encoded so that `R` handles them correctly. 
- Encoding guarantees that the computer will handle the special characters correctly.
- We will not spend much time on this; however, to read more about encoding text, [see this post.](https://cran.r-project.org/web/packages/utf8/vignettes/utf8.html)

**In general, make sure to switch back to English keyboard before coding!!!!!!**


---

# Data Types (Numeric)

Numeric data are "numbers"
- In a lot of programming languages, there is this distinction between floats and integers. This is also true in `R` but to a much smaller degree. 
- Unless explicitly stated as an integer, all "numbers" are numeric data.

--

```{r}
my_age    = 29
my_height = 5 + 11/12
c(my_age,my_height)
class(my_age)
class(my_height)
```

---

# Data Types (Numeric)

## Special Numeric Data

- There are a handful of special numeric values including Inf, -Inf and NaN
- Be careful when working with these

--

```{r}
1/0
log(0)
log(-1)
```

---


# Data Types (Numeric)

## Integers

"Whole numbers"
- Integers in R are distinguished from numeric data by having an L after the number part.
- The distinction between numeric and integers is not that important here in R.
  - We will mostly ignore the distinction.

--

```{r}
also_my_age = 29L
class(my_age)
class(also_my_age)
```

---

# Data Types (Logical)

Logical data are either `TRUE` or `FALSE`.
- `T` and `F` are equivalent.
- Not `true` and `false` though

--

```{r,error=TRUE}
R_is_fun  = TRUE
R_is_hard = FALSE
R_is_fun  == T
R_is_fun  == true
```

---

# Data Types (Logical)

## Operations

The most common use of logical data is for testing conditions and control flow.
- \>, <, >=, and <= test for greater/less than and/or equal to.
- & and | are the logical operators for "and" and "or" respectively.
  - Order of operations: & are evaluated before |.
- ! negates a logical: `!TRUE` becomes `FALSE` and vice versa,
- To test if equal, use two equal signs ==. Not equal !=.

---

# Data Types (Logical)

## Operations

--

```{r}
2 > 1
1 > 2 & 1 > 1/2
1 > 2 | 1 > 1/2
2 > 1 | 1 > 1/2
```

---

# Data Types (Logical)

## Operations

--

```{r}
1 > 1/2 & 1 > 2 | 3 > 2   #order of operations are important
(1 > 1/2) & (1 > 2) | (3 > 2)   #above is equivalent to this
1 > 2 & (1 > 1/2 | 3 > 2) #use parentheses to change the order of operation
0.5 == 1/2
3   != 2                  #that is ! followed by a =
```

---
# Data Types (Logical)

## Operations

%in% tests for containment

--

```{r}
R_is_fun
!R_is_fun
2 %in% c(1,2,3,4)
!(5 %in% c(1,2,3,4))
```

---
# Data Types (cont.)

## Special Logicals

- `NA` (Not Available) is a special type of logical data
  - Cannot be compared the same way.
  - `is.na` can detect `NA` and `NaN`
- `NaN` (Not a Number) is a special type of numeric data
  - Calculation result is not a number
  - `is.nan` only works for `NaN`
  
--

```{r}
NA + 1
is.na(NA)
is.nan(NaN)
```

---

# Data Types (Logical)

## Operations

If "mathematical operations" are performed on logicals, they will be coerced into 1s and 0s. 

--

```{r}
as.numeric(c(TRUE,FALSE))
TRUE + FALSE
sum(c(TRUE,FALSE,TRUE))
mean(c(TRUE,FALSE,TRUE))
```
---

# Data Types (Logical)

## Comparing Floats

Must be careful when testing for equality of floating point numbers ("decimals")

--

```{r}
0.1 + 0.2 == 0.3
```

What happened?
--

- Finite precision: floating point numbers aren't exact due to how numbers are stored in a computer.

```{r}
all.equal(0.1+0.2,0.3)
```

---
# Data Types (Logical)

You can read more about logical operators and types <a href="https://stat.ethz.ch/R-manual/R-devel/library/base/html/Logic.html" target="_blank">here</a> and <a href="https://stat.ethz.ch/R-manual/R-devel/library/base/html/logical.html" target="_blank">here</a>.

---
# Data Types (Complex)

Complex data are data that have complex numbers.
- We probably won't use these much but can pop-up if you're not careful.

```{r, error=TRUE}
2+3*sqrt(-1) #doesn't work
i            #doesn't work
2+3i         #works
class(2+3i)  #complex
```


---

# Assignment

In `R`, we can use either `<-` or `=` to handle assignment.<sup>1</sup>

.footnote[
<sup>1</sup> The `<-` is really a `<` followed by a `-`. It just looks like one thing b/c of the [font](https://github.com/tonsky/FiraCode) I'm using here.
]

--

### Assignment with `<-`

`<-` is normally read aloud as "gets". You can think of it as a (left-facing) arrow saying *assign in this direction*.

```{r}
a <- 10 + 5
a
```

--

Of course, an arrow can point in the other direction too (i.e. `->`). So, the following code chunk is equivalent to the previous one, although used much less frequently.
```{r}
10 + 5 -> a
```

---

# Assignment (cont.)

### Assignment with `=`

You can also use `=` for assignment.

```{r}
b = 10 + 10 ## Note that the assigned object *must* be on the left with "=".
b
```

--

### Which assignment operator to use?

Most R users (purists?) seem to prefer `<-` for assignment, since `=` also has specific role for evaluation *within* functions. 
- We'll see lots of examples of this later.
- But I don't think it matters; `=` is quicker to type and is more intuitive if you're coming from another programming language. (More discussion [here](https://github.com/Robinlovelace/geocompr/issues/319#issuecomment-427376764) and [here](https://www.separatinghyperplanes.com/2018/02/why-you-should-use-and-never.html).)
- Slides for this class uses `=`, but I personally prefer `<-`.

**Bottom line:** Use whichever you prefer. Just be consistent.

---
class: inverse, center, middle
name: datastructures

# Data Structures in R

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Data Structures

In base `R`, there are 5 main data structures (not exhaustive).

1. atomic vector
2. list
3. matrix/array
4. data.frame
5. factors

---
# Data Structures (Vectors)

- Vectors are a collection of multiple objects.
- There are two types of vectors:
  1. atomic vectors
  2. lists.
- Atomic vectors must be all of the same type of data.
- To create an atomic vector, put multiple objects within `c()` separated by commas.
  - **Never store an object as c! In general, avoid naming objects after existing functions**
  - Will break your code and it might take hours to figure out why.
- A list is a collection of atomic vectors.
- Atomic vectors are one dimensional

--

```{r}
numeric_grades = c(90,75,95,85,100,60,76)
letter_grades  = c("A-","C","A","B","A","D","C")
mixed_grades   = c("A", 95,"B",85,"C",75)
```

--

I said atomic vectors must be all of the same type. However, the last line ran without an error. What do you think happened?

---
# Data Structures (Vectors)

```{r}
class(numeric_grades)
class(letter_grades)
mixed_grades
class(mixed_grades)
```
---
# Data Structures (Vectors)

## Attributes

- Objects in R can have attributes. Each type of data structure (and objects more generally) will have different attributes.
- Types of attributes include (but not limited to):
 1. names
 2. dimnames
 3. class

---
# Data Structures (Vectors)

## Attributes

```{r} 
names(numeric_grades)
names(numeric_grades) = c("Student 1","Student 2","Student 3","Student 4",
                          "Student 5","Student 6","Student 7") #this is bad code!
names(numeric_grades)
```

--

```{r}
names(numeric_grades) = NULL
names(numeric_grades) = paste("Student",1:length(numeric_grades))
names(numeric_grades)
```
---

# Data Structures (Vectors)

## Indexing
- Indexing in `R` is very simple: it starts at 1 and count up by 1.
  - This is different than indexing in other programming languages which starts at 0.
  - Most "mathematical" languages will start at 1 e.g. `R`, `Matlab`, `Julia`.
- To index a vector, put `[i]` after the name of the vector where `i` is the ith position.

```{r}
some_numbers = c(27,22,94)
some_numbers[1]
some_numbers[3]
some_numbers[1:2]
```

---

# Data Structures (Vectors)

## Indexing
- Can use indexing to change values in an object   

--

```{r}
some_numbers[2]
some_numbers[2] = 23
some_numbers[2]
```

--

- Can sequentially index

--

```{r}
some_numbers[1:2][2]
```


---
# Data Structures (Vectors)

## Indexing

You can also use `c` to add object and `-` index to remove object

```{r}
some_numbers

c(some_numbers, 12)

some_numbers[-3]

some_numbers[-(2:3)]
```



---
# Data Structures (Lists)

- Lists are collections of atomic vectors.
- Since each atomic vector is still an atomic vector, within vector types have to be the same but across vectors can differ in type.

--

```{r}
names(letter_grades) = paste("Student",1:length(letter_grades))
grade_list = list(names(numeric_grades),numeric_grades,letter_grades)
grade_list
```

---

# Data Structures (Lists)

## Indexing
- Indexing for lists is almost the same as vectors, you just have another layer to deal with.
- To return an element of the list as a list, use single brackets: `[i]`
- To return an element of the list as a vector, use double brackets: `[[i]]`

--

```{r}
grade_list[1]
grade_list[[1]]
```

---
# Data Structures (Lists)

## Indexing
- Elements of a list can still have names

--

```{r}
names(grade_list) = c("student names","numeric grade","letter grade")
grade_list
```

---

# Data Structures (Lists)

## Indexing
- You can access an element of a list using it's name and the `$`

--

```{r}
grade_list$`numeric grade` #`` are used because of the space
```

--

- Double indexing still works

--

```{r}
grade_list[[1]][4]
grade_list$`student names`[4]
```

---

# Data Structures (Lists)

## Indexing

- You can add an element by assigning it to the list with `$` and remove an element by setting it to `NULL`


```{r}
grade_list$`Newfeature` <- 1:10

grade_list
```

---

# Data Structures (Lists)

## Indexing

- You can remove an element by setting it to `NULL`


```{r}
grade_list$`Newfeature` <- NULL

grade_list
```

---
# Data Structures (Matrix)

- Matrices are essentially two dimensional atomic vectors.
- Most things that apply to atomic vectors are true for matrices
- While we can make character matrices, they aren't very useful as matrices are most useful for mathematical operations.
- Due to the nature of this course, we will probably not use matrices much as we are not coding things by hand.

--
```{r}
num_mat = matrix(1:9,ncol=3)
num_mat
```

---
# Data Structures (Matrix)

## Indexing
- Indexing a matrix is similar to atomic vectors; two dimensions to index.
- If one dimension is omitted, returns that entire dimension.
- First index is for rows; second index is for columns.

```{r}
num_mat[1,2]
num_mat[1,]
num_mat[,2]
num_mat[5]
```

---
# Data Structures (Matrix)

## Indexing
- You can add a row using `rbind`, add a column using `cbind`.

```{r}
cbind(num_mat, 1:3)
rbind(num_mat, 1:3)
```

--

- You can still remove row/column using `-` (You cannot remove an single entry though).

```{r}
num_mat[,-2]

num_mat[-2,]
```

---

# Data Structures (Array)

- Arrays are 2 dimensional *or greater* matrices.
- Again, while arrays can be useful, we will likely not be working with them much.

```{r}
num_array = array(1:(3*3*2),dim = c(3,3,2))
num_array
```

---
# Data Structures (Array)

## Indexing

- Indexing is same as matrices, just the same number of indexes as dimensions
--

```{r}
num_array[1,2,1]
num_array[,,1]
num_array[1,2,]
num_array[12]
```

---
# Data Structures (Data.frame)

- data.frames are basically "data sets."
- Using established terminology, they are collections of atomic vectors in a rectangular format.
  - This may sound similar to lists; however, lists don't have a restriction on the geometric format.
    - Could have a lists of lists of lists etc.
- Each column of a data.frame must be of the same type.
- Each row is an observation.

```{r}
data(mtcars)
head(mtcars,5)
```

---
# Data Structures (Data.frame)

- `names()` on a data.frame returns the names of the variables.
- `length()` and `ncol()` return the number of variables in the data.frame
- `nrow()` returns the number of rows in the data.frame.
- Note: if you don't use `head()`, `tail()` or another way to restrict which rows are returned, returning a data.frame will return all rows, which is annoying.
 - This is fixed in a popular library called `data.table` that we will learn later.
 - This is also fixed in tibble, a modern reimagining of the data frame, in `tidyverse`.

```{r}
mtcars
```
---
# Data Structures (Data.frame)

## Indexing
- Indexing a data.frame is similar to indexing a matrix; the columns can have names

```{r}
mtcars[1:5,"mpg"]
mtcars$mpg[1:5]
mtcars[1:5,1]
row.names(mtcars)[1:5]
```

---
# Data Structures (Data.frame)

## Indexing
- Adding/removing row/column is also similar to matrix using `rbind`/`cbind`/`-`. Also, we can use `$` or assing to `NULL` similar to list.


```{r, eval=FALSE}
mtcars[-1,]
mtcars[,-1]
mtcars$NewVar <- 1
mtcars$NewVar <- NULL

# Need same number of columns when doing rbind
rbind(mtcars, 1:11)

# Need same number of rows when doing cbind
cbind(mtcars, 1:32)
```

---
# Data Structures (Factors)

- Factors are essentially categorical variables.
- Factors can organized text into different groups.
- Are only really useful to data analysis and modeling.
- Factors have different "levels" which are the groups.
- Can be tricky when first learning, but ultimately rather simple.

```{r factors}
# Draw 10 ramdom letters with replacement
fact_groups = letters[sample(1:26,10,replace=T)]
fact_groups
# Form the factor group
fact_groups = factor(fact_groups,levels=letters)
fact_groups
```

---
class: inverse, center, middle
name: help

# Getting help

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>


---

# Help

For more information on a (named) function or object in R, consult the "help" documentation. For example:

```R
help(plot)
```

Or, more simply, just use `?`:
  
```R
# This is what most people use.
?plot 
```

--

**Aside:** See the *Examples* section at the bottom of the help file?
- You can run them with the `example()` function. Try it: `example(plot)`.

---

# Help (cont.)

### Vignettes

For many packages, you can also try the `vignette()` function, which will provide an introduction to a package and it's purpose through a series of helpful examples.
- Try running `vignette("dplyr")` in your console now.

--

I highly encourage reading package vignettes if they are available. 
- They are often the best way to learn how to use a package.

--

One complication is that you need to know the exact name of the package vignette(s).
- E.g. The `dplyr` package actually has several vignettes associated with it: "dplyr", "window-functions", "programming", etc.
- You can run `vignette()` (i.e. without any arguments) to list the available vignettes of every *installed* package installed on your system. 
- Or, run `vignette(all = FALSE)` if you only want to see the vignettes of any *loaded* packages.

---

# Help (cont.)

### Demos

Similar to vignettes, many packages come with built-in, interactive demos.

To list all available demos on your system:<sup>1</sup> 

```r
demo(package = .packages(all.available = TRUE))
```

.footnote[
<sup>1</sup> How would you limit the demos to one particular package?
]

--

To run a specific demo, just tell R which one and the name of the parent package. For example:

```r
demo("graphics", package = "graphics")
```

---

# HW 1

I will release HW 1 on this afternoon

It will be due on Tuesday, 9/11, before our class begins. 

...*because* this is introductory material, I am going to ask that you ***do not*** use LLMs/AI models for help with this homework specifically.

I know I said you could (and should!) use them for our class. And you should! But you need to build strong coding foundations first

</br>

Recommended order of resources:

Try Yourself &rarr; Slides &rarr; Online Threads &rarr; Classmates &rarr; ChatGPT (but ask for explaination)

---

class: inverse, center, middle

# Next lecture(s): Objects and the OOP approach

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>



```{r gen_pdf, include = FALSE, cache = FALSE, eval = TRUE}
infile = list.files(pattern = '.html')
#pagedown::chrome_print(input = infile, timeout = 100000)
```
