---
title: "ECON 370: Assignment 5"
author: "Jiaxi Li"
date: "Due: Before class on December 2nd, 2025"
output: 
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{geometry}
  - \usepackage{float}
  - \usepackage{graphicx}
  - \usepackage{caption}
  - \usepackage{multirow}
  - \usepackage{array}
  - \usepackage[affil-it]{authblk}
  - \usepackage{lscape}
  - \usepackage[bottom]{footmisc}
  - \usepackage{amssymb}
  - \usepackage{amsmath}
  - \geometry{top=1in, bottom=1in}
  - \setlength{\parindent}{0pt}
  - \setlength{\parskip}{12pt}
---

```{r setup, include=FALSE}
# Note, this is global setting for all code chunks
knitr::opts_chunk$set(eval = FALSE, include = FALSE)
```

Names of all group members:

# Directions

## LLMs and AI tools may be used for this assignment, but only with careful verification of their output.

Use the Rmd template and write code to answer the following questions. In order to ensure smooth grading, I will ask you to write both your code and explanation (as comment) in code chunk. Please write your code and answer in the corresponding place.

In this homework, we will utilize simulation and test statistics to answer some theoretical questions and basic complex models.

Each question is worth three points. For each section, a "question" corresponds to a number.

Please **comment** your code clearly. 0.5 point per question will be deducted if there is a significantly lack of commenting. Also, the easier your code is to read, the more likely I will be able to figure out what you're doing.

For any process requires randomness, you need to **set seed to 2025**. For for loops, you need to **preallocate** the result. Failing to do so will result in 0.5 point per question deduction.

Remember to ***run Section O*** before you proceed. After finishing all code, make sure to **restart** R session and **clear** everything from your workspace. **Rerun** everything to ensure there is no dependency issue in your code.

For ***Section III***, I will let you modify some parts of the code labeled as **?**. The major goal is for you to compare the results.

You may work in groups of three to four people for this homework. Please submit one script per group and clearly indicate who was in your group for the assignment.

------------------------------------------------------------------------

# O. Data

***This is the code to extract data online and load necessary packages. Do not modify. Run this section before you proceed.***

Similar to HW4, we are also creating some new data based on this formula:

$$
Y = 0.5 X_1 + 1 X_2 + 1.5 X_3 + 2 X_4 + e
$$

We would also have $X_1$ to be positively correlated to $X_2, X_5$ and negatively correlated to $X_3$. $X_1$ is not correlated with $X_4$ at all. There are 476 variables have no contribution to $Y$. We are also creating an $Y_2$ to be created by adding random coefficients for all the 480 $X$ variables, saved as "**X_alt**". The actual coefficients are saved as "**b_actual**".

```{r}
# Data Loading in Section, do not modify
library(tidyverse) # ggplot2 is already in tidyverse
library(glmnet) # for many other regressions such as: Ridge, LASSO, logistic, etc.
library(ranger) # for Random Forest
library(xgboost) # for Boosting
library(caret) # for CV

options(torch.threshold_call_gc = 4000) # garbage collector for torch
library(torch)        # Core PyTorch in R
library(luz)          # High-level interface for torch
library(torchvision)  # Useful for datasets and transformations

# Simulate data based on the model
set.seed(2025)
# Set observation size
Nobs = 1000
Noise = 3*rnorm(Nobs)
X = matrix(rnorm(5*Nobs), ncol = 5)
# Make X_2,X_5 to be positively correlated to X_1
X[,2] = 0.6*X[,1] + sqrt(1 - 0.6^2) * X[,2]
X[,5] = 0.6*X[,1] + sqrt(1 - 0.6^2) * X[,5]
# Make X_3 to be negative correlated to X_1
X[,3] = -0.6*X[,1] + sqrt(1 - 0.6^2) * X[,3]
# Create the Y variable
Y = 0.5*X[,1] + 1*X[,2] + 1.5*X[,3] + 2*X[,4] + Noise

# Create alternative variables who are just random noise
X_alt = cbind(X, matrix(rnorm(475*Nobs), ncol = 475))

# Add names to X
colnames(X) = paste0("X", 1:5)
# Add names to X_alt
colnames(X_alt) = paste0("X", 1:480)

# Create b_actual
b_actual = sort(rnorm(480))*0.05

# Create Y_2 based on all 1080 variables
Y_2 = rowSums(sweep(X_alt, 2, b_actual, "*")) + Noise

# Remove the random noise
rm(Noise)

# id for testing
testing_id <- sample(1:nrow(X_alt), floor(nrow(X_alt)*0.3))
X_alt_train <- X_alt[-testing_id,]
X_alt_test <- X_alt[testing_id,]
Y_train <- Y[-testing_id]
Y_test <- Y[testing_id]
Y_2_train <- Y_2[-testing_id]
Y_2_test <- Y_2[testing_id]
```

------------------------------------------------------------------------

# I. P-Hacking for Sparse Data

We learned about p-value in *lecture 10* which can determine variable statistical significance.

To find significant results, some researchers selectively report results, run many different tests, or stopping data collection once a desired result is found. This is called **p-hacking** or **data dredging**. Let's check if p-hacking is a good idea with simulations.

1.  We have created 480 variables in "X_alt", how many are truly related to $Y$? Given that p-value is the probability that an **irrelevant** variable can be found more extreme than the given variable, with $\alpha = 0.05$, what is the chance that an **irrelevant** variable gets falsely concluded as statistically significant? Given that there are for $Y$, how many **irrelevant** variables would you expect to see have a low p-value?

```{r}
# Answer:
# # of variables truly relate to Y: 
# chance that an irrelevant variable gets falsely concluded as statistically significant: 
# how many irrelevant variables would you expect to see have a low p-value: 
```

2.  Now, let's use **for loop** to run **regressions**, **one variable at a time without intercept**, and save the p-values in "**Result**". Check the proportion and number of **irrelevant** variables with p\<0.05. Is the proportion similar to your guess in 1?

```{r}
# Code Below, remember commenting

```

```{r}
# Answer:
# whether proportion is similar to your guess in 1: 
```

3.  If I am a researcher, who searched online and randomly (without reasoning) a list of 280 variables. I run a regression for each variable to find the useful ones. I am really "conservative" in the sense that I use $\alpha = 0.01$ and concludes that there are about **8** useful variables for $Y$ prediction. Am I a good researcher?

```{r}
# Answer: 
```

4.  Let's use the idea of **training and testing** data to split the sample (I have already split them for you). Regress Y against **each 476 irrelevant variables** in the training data "X_alt_train" and "Y_train". Find the variables significant (p \< 0.05) and run those variables again in the testing data "X_alt_test" and "Y_test". How many variables are still significant? Does OOS test help to flag false signals? Does OOS test eliminate "luck" (in the sense a lucky irrelevant variable is concluded as significant)?

Hint: Your should skip the first 4 relevant variables when calculating the total number and proportion of significant but irrelevant variables.

```{r}
# Code Below, remember commenting

```

```{r}
# Answer:
# Does OOS test help to flag false signals? 
# Does OOS test eliminate "luck"? 
```

5.  We have seen the p-value conclusion is problematic, let's examine OOS predictability. Take **only** the **statistically significant** variables ($p<0.05$) from your training set, run a multiple regression in-sample (using **training set** and all "significant" variables together in **one** regression without intercept, call this "Model2"), and then use this regression to fit the test set with "**predict**" function and calculate the OOS $R^2$. Try the same for just the first 4 relevant variables (call this "Model1") and calculate the OOS $R^2$ again. How good are the predictions if you include irrelevant but significant variables?

Hint: For `lm` model and associated predict, you need to input a data.frame (or tibble or data.table).

```{r}
# Code Below, remember commenting


# Answer: 
```

6.  Let's try to use all variables in regression now and examine OOS performance. TRun a multiple regression in-sample (using **training set** and all variables together in **one** regression without intercept, call this "**Model3**"), and then use this regression to fit the test set with "**predict**" function and calculate the OOS $R^2$. How good are the predictions if you just use everything without careful reasoning?

```{r}
# Code Below, remember commenting


# Answer: 
```

7.  Let's try the LASSO (**Model5**) and Ridge (**Model4**) with cross validation. Since this is a cross-sectional data, you can just use "**cv.glmnet**" to find the best model. (You can check lec 10 template code, remember to set seed to 2025.) Estimate the OOS $R^2$ for both models. Which method does better?

Hint: To fit new data with `cv.glmnet`, you should use `predict(model, newx = new_data, s = "lambda.min")`. Note for `glmnet`, you should use matrix.

```{r}
# Code Below, remember commenting


# Answer: 
```

8.  Why do you think the Ridge and LASSO have such performance for $Y$? (Hint: related to HW4 idea.)

```{r}
# Answer: 
```

------------------------------------------------------------------------

# II. P-Hacking for Dense Data

Now, let's repeat the exact exercise for Dense Data. Many hint can be found in the previous section.

1.  We have created 480 variables in "X_alt", how many are truly related to **Y_2**?

```{r}
# Answer:
# # of variables truly relate to Y_2: 
```

2.  Now, let's use **for loop** to run **regressions**, **one variable at a time without intercept**, and save the p-values in "**Result_alt**". Check the proportion and number of **significant** variables with p\<0.05. Why do you think there are so few significant variables?

```{r}
# Code Below, remember commenting

```

```{r}
# Answer: 
```

3.  If I am a researcher, who searched online and randomly (without reasoning) a list of 480 variables. I run a regression for each variable to find the useful ones. I am really "conservative" in the sense that I use $\alpha = 0.01$ and concludes that there are about **12** useful variables for Y_2 prediction. Am I a good researcher?

```{r}
# Answer: 
```

4.  Let's use the idea of **training and testing** data to split the sample (I have already split them for you). Regress Y_2 against **all 480 variables** in the training data "X_alt_train" and "Y_2_train". Find the variables significant (p \< 0.05) and run those variables again in the testing data "X_alt_test" and "Y_2_test". How many variables are still significant? Does OOS test confirm their relevance? Why?

```{r}
# Code Below, remember commenting

```

```{r}
# Answer: 
```

5.  Let's try to use all variables in regression now and examine OOS performance. Run a multiple regression in-sample (using **training set** and all variables together in **one** regression without intercept, call this "**Model6**"), and then use this regression to fit the test set with "**predict**" function and calculate the OOS $R^2$. How good are the predictions if you just use everything here? Guess why?

```{r}
# Code Below, remember commenting


# Answer: 
```

6.  Let's try the LASSO (**Model7**) and Ridge (**Model8**) again with cross validation ("**cv.glmnet**", remember to set seed to 2025). Estimate the OOS $R^2$ for both models. Which method does better?

```{r}
# Code Below, remember commenting


# Answer: 
```

7.  Why do you think the Ridge and LASSO have such performance for Y_2?

```{r}
# Answer: 
```

------------------------------------------------------------------------

# III. Regression Trees and Neural Networks

In this section, we will try to modify **lecture 11 template code** in class for the tree models. I will provide you with analysis code about neural networks and I want you to make sense of the code and the result.

Warning: CV is much easier for cross-sectional data here. You probably need to use loop when you work with time-series.

1.  Let's just use all variables in X_alt to predict **Y**. Let's use `ranger` with `caret` for Random Forest + 5-fold cross validation for Y_train. Name this model "**Model9**". Fit the X_alt_test to Model9, calculate and report the OOS $R^2$. How well is the Random Forest Performance?

Note: CV takes a long time.

```{r}
# Modify Code Below, replace the ? with your code
# Make data a data.frame
train_df <- data.frame(Y = Y_train, X_alt_train)

set.seed(2025)
# Define train control with 5-fold CV
train_ctrl <- trainControl(method = "?", number = 5)

# Train random forest using ranger via caret
Model9 <- train(
  ?,
  data = train_df,
  method = "ranger",
  trControl = train_ctrl,
  tuneLength = 3 # for better results, this should be 10
)

# We can directly fit new data to rf
Fitted9 <- ?(Model9, newdata = X_alt_test)

# Estimate the OOS R^2
R_sq_OOS9 = ?

paste0("The out-of-sample R^2 for Random Forest with Y is: ", round(R_sq_OOS9, 4))
```

```{r}
# Answer: 
```

2.  Repeat with Boosting with `xgboost` to fit **Y**, use the `xgb.cv` function for cross validation with early stopping and refit the model. Please use the exact code but change to different data. Fit the X_alt_test to **Model10**, calculate and report the OOS $R^2$. How well is the Boosting Performance?

Hint: You need to use matrix here for `xgb.cv`. Set seed twice as 2025 for CV and refitting.

Note: CV takes a long time.

```{r}
# Modify Code Below, replace the ? with your code
set.seed(2025)
# 5-fold CV for boosting
cv_res10 <- ?(
  data = X_alt_train,
  label = Y_train,
  nrounds = 2000,
  eta = 0.01,
  nfold = 5,
  objective = "reg:squarederror",
  early_stopping_rounds = 50,
  verbose = 0
)

# We have to refit the boosted tree though
set.seed(2025)
Model10 <- ?(
  data = X_alt_train,
  label = Y_train,
  objective = "reg:squarederror",
  nrounds = cv_res10$best_iteration,
  eta = 0.01,
  verbose = 0   # silence the fitting message
)

# fit new data to xgb_model
Fitted10 <- ?(Model10, newdata = X_alt_test)

# Estimate the OOS R^2
R_sq_OOS10 = ?

paste0("The out-of-sample R^2 for Boosting with Y is: ", round(R_sq_OOS10, 4))
```

```{r}
# Answer: 
```

3.  Here is the provided Neural Networks model for **Y**. How many layers do we have? How many parameters in total do we have? How does the model work based on OOS $R^2$?

```{r}
# Just run the code below
# Set random seed for reproducibility
torch_manual_seed(2025)

# Define single-hidden-layer NN module
modnn <- nn_module(
  initialize = function(input_size) {
    # 1st Hidden layer with 128 neurons
    self$hidden1 <- nn_linear(input_size, 128)
    
    # 2nd Hidden layer with 64 neurons
    self$hidden2 <- nn_linear(128, 64)
    
    # 3rd Hidden layer with 32 neurons
    self$hidden3 <- nn_linear(64, 32)
    
    # 4th Hidden layer with 16 neurons
    self$hidden4 <- nn_linear(32, 16)
    
    # Activation function
    self$activation <- nn_relu()
    
    # Dropout for regularization (40% dropout)
    self$dropout <- nn_dropout(0.4)
    
    # Output layer (regression output)
    self$output <- nn_linear(16, 1)
  },
  forward = function(x) {
    x |>
      self$hidden1() |>     # Linear transformation
      self$activation() |>  # Apply ReLU
      self$dropout() |>     # Apply dropout
      self$hidden2() |>     # Linear transformation
      self$activation() |>  # Apply ReLU
      self$dropout() |>     # Apply dropout
      self$hidden3() |>     # Linear transformation
      self$activation() |>  # Apply ReLU
      self$dropout() |>     # Apply dropout
      self$hidden4() |>     # Linear transformation
      self$activation() |>  # Apply ReLU
      self$dropout() |>     # Apply dropout
      self$output()         # Final output
  }
)

# Skip the scaling since our data is already standard normal. It is better to scale for real data.

set.seed(2025)
# Find val_id for validation set
val_id <- sample(1:nrow(X_alt_train), floor(nrow(X_alt_train)*2/7))

# Set up model with luz
modnn <- modnn |>
  setup(
    loss = nn_mse_loss(),       # Regression MSE loss
    optimizer = optim_adam,
    metrics = list(luz_metric_mae())
  ) |>
  set_hparams(
    input_size = ncol(X_alt_train)   # Specify input size
  )

# Fit the model
fitted <- modnn |>
  fit(
    data = list(X_alt_train[-val_id,], matrix(Y_train[-val_id], ncol = 1)), # training
    valid_data = list(X_alt_train[val_id,], matrix(Y_train[val_id], ncol = 1)), # validation
    epochs = 80,
    # call back for early stopping
    callbacks = list(
      luz_callback_early_stopping(
        monitor = "valid_loss",
        patience = 15
      )
    )
  )
```

```{r}
# Show the model structure
fitted$model
```

```{r}
# Plot training metrics
plot(fitted)
```

```{r}
# Find the predicted with fitted model
Fitted11 <- predict(fitted, X_alt_test) |>
  as.numeric()

# Estimate the OOS R^2
R_sq_OOS11 = 1 - var(Fitted11 - Y_test)/var(Y_test)

paste0("The out-of-sample R^2 for Nerual Networks with Y is: ", round(R_sq_OOS11, 4))
```

```{r}
# Answer:
# # of layers: 
# # of parameters in total: 
# OOS R^2: 
```

4.  Let's just use all variables in X_alt to predict **Y_2**. Let's use `ranger` with `caret` for Random Forest + 5-fold cross validation for Y_2_train. Name this model "**Model12**". Fit the X_alt_test to Model11, calculate and report the OOS $R^2$. How well is the Random Forest Performance for Y_2?

Note: CV takes a long time.

```{r}
# Modify Code Below, replace the ? with your code
# Make data a data.frame
train_df2 <- data.frame(Y = Y_2_train, X_alt_train)

set.seed(2025)
# Define train control with 5-fold CV
train_ctrl <- trainControl(method = "?", number = 5)

# Train random forest using ranger via caret
Model12 <- train(
  ?,
  data = train_df2,
  method = "ranger",
  trControl = train_ctrl,
  tuneLength = 3 # for better results, this should be 10
)

# We can directly fit new data to rf
Fitted12 <- ?(Model12, newdata = X_alt_test)

# Estimate the OOS R^2
R_sq_OOS12 = ?

paste0("The out-of-sample R^2 for Random Forest with Y_2 is: ", round(R_sq_OOS12, 4))
```

```{r}
# Answer: ?
```

5.  Repeat with Boosting with `xgboost` to fit **Y_2**, use the `xgb.cv` function for cross validation with early stopping and refit the model. Please use the exact code but change to different data. Fit the X_alt_test to **Model13**, calculate and report the OOS $R^2$. How well is the Boosting Performance?

Hint: You need to use matrix here for `xgb.cv`. Set seed twice as 2025 for CV and refitting.

```{r}
# Modify Code Below, replace the ? with your code
set.seed(2025)
# 5-fold CV for boosting
cv_res13 <- ?(
  data = X_alt_train,
  label = Y_2_train,
  nrounds = 2000,
  eta = 0.01,
  nfold = 5,
  objective = "reg:squarederror",
  early_stopping_rounds = 50,
  verbose = 0
)

# We have to refit the boosted tree though
set.seed(2025)
Model13 <- ?(
  data = X_alt_train,
  label = Y_2_train,
  objective = "reg:squarederror",
  nrounds = cv_res13$best_iteration,
  eta = 0.01,
  verbose = 0   # silence the fitting message
)

# fit new data to xgb_model
Fitted13 <- ?(Model13, newdata = X_alt_test)

# Estimate the OOS R^2
R_sq_OOS13 = ?

paste0("The out-of-sample R^2 for Boosting with Y_2 is: ", round(R_sq_OOS13, 4))
```

```{r}
# Answer: 
```

6.  Here is the provided Neural Networks model for **Y_2**. How many layers do we have? How many parameters in total do we have? How does the model work based on OOS $R^2$?

General idea for dense data with small coefficient would be to fit a neural networks model with one layer and small number of neurons (shallow & small). Twisting an Neural Networks Model to fit data is almost an art.

```{r}
# Just run the code below
# Set random seed for reproducibility
torch_manual_seed(2025)

# Define single-hidden-layer NN module
modnn <- nn_module(
  initialize = function(input_size) {
    # 1st Hidden layer with 16 neurons
    self$hidden1 <- nn_linear(input_size, 16)
    
    # Activation function
    self$activation <- nn_relu()
    
    # Dropout for regularization (40% dropout)
    self$dropout <- nn_dropout(0.4)
    
    # Output layer (regression output)
    self$output <- nn_linear(16, 1)
    
    # ---- CUSTOM SMALL INITIALIZATION ----
    nn_init_normal_(self$hidden1$weight, mean = 0, std = 0.005)
    nn_init_constant_(self$hidden1$bias, 0)

    nn_init_normal_(self$output$weight, mean = 0, std = 0.005)
    nn_init_constant_(self$output$bias, 0)
  },
  forward = function(x) {
    x |>
      self$hidden1() |>     # Linear transformation
      self$activation() |>  # Apply ReLU
      self$dropout() |>     # Apply dropout
      self$output()         # Final output
  }
)

# Skip the scaling since our data is already standard normal. It is better to scale for real data.

set.seed(2025)
# Find val_id for validation set
val_id <- sample(1:nrow(X_alt_train), floor(nrow(X_alt_train)*2/7))

# Set up model with luz
modnn <- modnn |>
  setup(
    loss = nn_mse_loss(),       # Regression MSE loss
    optimizer = optim_adam,
    metrics = list(luz_metric_mae())
  ) |>
  set_hparams(
    input_size = ncol(X_alt_train)   # Specify input size
  ) |>
  set_opt_hparams(
    lr = 1e-4
  )

# Fit the model
fitted <- modnn |>
  fit(
    data = list(X_alt_train[-val_id,], matrix(Y_2_train[-val_id], ncol = 1)), # training
    valid_data = list(X_alt_train[val_id,], matrix(Y_2_train[val_id], ncol = 1)), # validation
    epochs = 400,
    # call back for early stopping
    callbacks = list(
      luz_callback_early_stopping(
        monitor = "valid_loss",
        patience = 15
      )
    )
  )
```

```{r}
# Show the model structure
fitted$model
```

```{r}
# Plot training metrics
plot(fitted)
```

```{r}
# Find the predicted with fitted model
Fitted14 <- predict(fitted, X_alt_test) |>
  as.numeric()

# Estimate the OOS R^2
R_sq_OOS14 = 1 - var(Fitted14 - Y_2_test)/var(Y_2_test)

paste0("The out-of-sample R^2 for Nerual Networks with Y_2 is: ", round(R_sq_OOS14, 6))
```

```{r}
# Answer:
# # of layers: ?
# # of parameters in total: ?
# OOS R^2: ?
```

7.  Run the following code. I am calling it "repeated" here but that is not the official word for the idea. What single word on Lecture 11 would describe what I am doing here? How does the performance change?

```{r}
# Just run the code below
# repeated fit the NN Model
n_repeated <- 5
repeated_models <- vector("list", n_repeated)

for (i in 1:n_repeated) {
  cat("Training model", i, "\n")
  torch_manual_seed(123 + i)   # different seeds for diversity
  
  set.seed(2025 + i)
  # Find val_id for validation set
  val_id <- sample(1:nrow(X_alt_train), floor(nrow(X_alt_train)*2/7))


  # Fit the model
  repeated_models[[i]] <- modnn |>
  fit(
    data = list(X_alt_train[-val_id,], matrix(Y_2_train[-val_id], ncol = 1)), # training
    valid_data = list(X_alt_train[val_id,], matrix(Y_2_train[val_id], ncol = 1)), # validation
    epochs = 400,
    # call back for early stopping
    callbacks = list(
      luz_callback_early_stopping(
        monitor = "valid_loss",
        patience = 15
      )
    )
  )

}

# Make predictions for each repeated model
preds_list <- lapply(repeated_models, function(model) {
  as.numeric(predict(model, X_alt_test))
})

# Average predictions
preds_repeated <- Reduce(`+`, preds_list) / n_repeated

# Estimate the OOS R^2
R_sq_OOS_repeated = 1 - var(preds_repeated - Y_2_test)/var(Y_2_test)
```

```{r}
paste0("The out-of-sample R^2 for repeated Nerual Networks with Y_2 is: ", round(R_sq_OOS_repeated, 6))
```

```{r}
# Answer:
# One word: 
# Performance: 
```

8.  What is your conclusion about overall OOS results for Y and Y_2? Does complicated model always work better than simple models? Why? (Hint: What is the DGP for Y and Y_2?)

```{r}
# Answer: 
```
