---
title: "Data Science for Economists"
# subtitle: "<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>"
subtitle: "Example Presentation"
author: "Jiaxi Li"
date: "University of North Carolina | [ECON 370](https://github.com/JiaxiLi1995/ECON370)" #"`r format(Sys.time(), '%d %B %Y')`"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
  fig.align="center", 
  fig.height=5, fig.width=10, 
  # out.width="748px", #out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=F, echo=F, warning=F, message=F
  )
library(fontawesome)
library(microbenchmark)
library(dplyr)
library(tidyr)
library(ggplot2)
set.seed(123)

# This function is used to color text
color_text <- function(text, color){
  if (knitr::is_html_output()) {
    paste0('<span style="color:', color, '">', text, '</span>')
  } else {
    paste0('\\textcolor{', color, '}{', text, '}')
  }
}
```

# Introduction

In this section, you would ***briefly*** summarize the highlights of your project. Think that this is a project report to your boss. Convince that person why your model and strategy is worth to invest in and this slide is to catch your boss's eyes!

An example:

**Hightlight of the project**:
- Data: 5 Macro predictors, 5 financial market indicators, 1 geopolitical indicator.
- Model: *Linear Model* as benchmark and *Neural Network Model* as the advanced model
- Achievement: With the more complex model, I am able to move the predictive $R^2$ from `r color_text("XX", "blue")` to `r color_text("XX", "red")`. With a strategy of XXX, I was able to boost the overall return in the period XXX from XX to XX.

---
# Data

Here is a table of data that I am using with brief reasoning:


### Table: Possible Data Related to Stock Returns

| Variable | Type | Reason |
|----------------|------------|-----------------------|
| GDP | Macro | Reflects economic growth |
| Lag Return | Financial | Captures return persistence |
| ... | ... | ... |
| ... | ... | ... |

--

I would particularly point out XX variable. Although it is not standard, due to XXX, I believe that it has value to predict returns.

---
# Model

Beyond linear model, I also used XX model for the analysis.

XX model is a XX model and it uses XX technique to get a better fit of the data.

There are some considerations of the XX model and it uses XX as tuning parameters.


---
# Main Result

With an expanding window from XX to XX date, and XX model, here is the out of sample $R^2$ and the actual vs. fitted returns plot:

(Here, you can save the plot from your main script, upload it onto GitHub, and then load the data here. I will demonstrate using this method.)

```{r, message = F}
library(ggplot2)
# Put your mouse on "Raw" and right click, then "copy link address"
Testing_result <- readRDS(url("https://github.com/JiaxiLi1995/Final_proj_data/raw/refs/heads/main/Testing_result.rds"))

Testing_result |>
  ggplot(aes(actual_return, fitted_return)) +
  geom_point() + 
  geom_smooth(method = "lm", se = TRUE, color = "red")  # Fitted regression line
```



---
# Other Results (name your own title)

Here, you can compare different trading strategies, explore the strong point and weak point of your model etc... I will show the result of trading strategies here:

With the buy/hold when there is positive predicted return and sell/stay out otherwise, the performance of the XX model is better than the linear model:

```{r}
Trading_Strategy <- readRDS(url("https://github.com/JiaxiLi1995/Final_proj_data/raw/refs/heads/main/Trading_Strategy.rds"))

Trading_Strategy |>
  pivot_longer(c(buyLMpositive, buyANDhold), names_to = "Strategy", values_to = "CR") |>
  ggplot(aes(predicting_date, CR, color = Strategy)) +
  geom_line(linewidth = 1) +
  theme_minimal() +
  labs(
    title = "Cumulative Returns with Different Trading Strategies",
    x = "Date",
    y = "Cumulative Returns",
    color = "Trading Stategy"
  ) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5)  # <-- Center the title
  )


```

---

class: inverse, center, middle


### Disclaimer: I have used ChatGPT to help me create tables in Rmd and making some of my code more efficient.


<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>


```{r gen_pdf, include = FALSE, cache = FALSE, eval = TRUE}
# Here should be the correct html file name
infile = "Example_presentation.html"
#pagedown::chrome_print(input = infile, timeout = 1000)
```
